---
title: "Raw Data & Initial Cleaning"
format:
  html: 
    toc: true
    toc-depth: 6
    embed-resources: true
output: html_document
editor_options: 
  chunk_output_type: console
---

In stage 1, we import the raw dataset and make any necessary revisions to variable names. 

## Preparations
### Packages
```{r}
pacman::p_load(tidyverse, haven, here, memisc)
```

## Import Dataset
```{r}
p1_data <- read_sav(Sys.getenv("RAW_PATH"))
```

## Codebook
We will generate a .txt file codebook. 

```{r}
#| eval: FALSE
Write(codebook(p1_data),
      file = "output_txt/raw_codebook.txt")
```

You can find the raw codebook in the output_txt folder. 

## Lowercase All Variable Names

```{r}
names(p1_data) <- tolower(names(p1_data))
```

## Removing Duplicates

Here we will remove all resposnes that are exactly the same, which can be **truly** consider duplicates. 

```{r}
duplicated(p1_data)
```

Here we can see that there are no direct duplicates. So we will keep all participants here.

## Creating an ID variable

```{r}
p1_data$internal_id <- 1:(nrow(p1_data))

head(p1_data$internal_id)
```

We create our own ID variable, but in this case we just made it correspond to their row number.

## Save Results
```{r}
saveRDS(p1_data, "work_data/p1_data.Rds")
```