---
title: "Ego-Alter Variables"
format:
  html: 
    toc: true
    toc-depth: 3
    embed-resources: true
output: html_document
editor_options: 
  chunk_output_type: console
---
:::callout-note
**Overall Summary**: This phase of data cleaning will

- prepare data for creating ego-network datasets
- correct alter variable names
- correct alter values
:::
## Preparations
### Packages
```{r}
pacman::p_load(tidyverse, haven, here, labelled, egor)
```
## Import Dataset
```{r}
p3_data <- readRDS("work_data/p2_data.Rds")
```

## Prepare Data
### Create EgoID variable
```{r}
p3_data <- p3_data |>
  mutate(egoid = 1:n()) |>
  select(egoid, everything())
```

### Remove All Text Variables
EgoR package cannot create an ego network data if text variables/values exist. Thus if any variables contain text, we will drop.

```{r}
p3_data <- p3_data |> 
  select(-contains("TEXT"))
```

## Correct Errors Variable Names
### Alter Service Attendance 
It looks like two variables were named incorrectly. The variable to measure service attendance of an alter is typically `alt_ra_X` where X = the number of the alter.
Here's what that variable label looks like:

```{r}
get_variable_labels(p3_data$alt_ra_1)
```

However, there are two variables with **incorrect** variable names. 

```{r}
get_variable_labels(p3_data$alr_ra_2)
get_variable_labels(p3_data$alr_ra_3)
```

Here, it's clear that the variables measure the same construct, yet they are named incorrectly. Let's fix that.

```{r}
p3_data <- rename(p3_data,
                    alt_ra_2 = alr_ra_2,
                    alt_ra_3 = alr_ra_3)
```

## Recode Values
### Alter-Alter Ties 

The value labels are coded in a manner that will bias our analyses later. 

```{r}
get_variable_labels(p3_data$ap1_2)
get_value_labels(p3_data$ap1_2)
```

As you can see, `No = 2` while `Yes = 1`. However, when we calculate the sum, we want to make sure that we only count the `Yes`'s, thus it makes more sense to code No as 0, and Yes as 1. 

```{r}
p3_data <- p3_data |>
  mutate(across(contains("ap1_") | contains("ap2_") | contains("ap3_") | contains("ap4_") | contains("ap5_") | contains("ap6_") | contains("ap7_") | contains("ap8_") | contains("ap9_"), ~ifelse(. == 2, 0, .)))
```

The code above specifies: `If value is equal to 2, then code as 0. If the condition is not true, leave as is`.

### Religious Importance
```{r}
get_variable_labels(p3_data$alt_ri_1)
get_value_labels(p3_data$alt_ri_1)
table(p3_data$alt_ri_1)
```

Similar to above, we want to make `No` = 0, and `Yes` = 2, so we simply subtract 1 from all values. 

```{r}
p3_data <- p3_data |>
  mutate(across(contains("alt_ri_"), ~. - 1))
table(p3_data$alt_ri_1)
```

We then need to correct the value labels.

```{r}
p3_data <- set_value_labels(
  p3_data,
  alt_ri_1 = c(No = 0, Yes = 1),
  alt_ri_2 = c(No = 0, Yes = 1),
  alt_ri_3 = c(No = 0, Yes = 1),
  alt_ri_4 = c(No = 0, Yes = 1),
  alt_ri_5 = c(No = 0, Yes = 1),
  alt_ri_6 = c(No = 0, Yes = 1),
  alt_ri_7 = c(No = 0, Yes = 1),
  alt_ri_8 = c(No = 0, Yes = 1),
  alt_ri_9 = c(No = 0, Yes = 1),
  alt_ri_10 = c(No = 0, Yes = 1)
)
```

Let's check the updated value labels.

```{r}
get_value_labels(p3_data$alt_ri_1)
```

### Religious Attendance

We need to recode values as missing for analytic purposes.

```{r}
get_value_labels(p3_data$alt_ra_1)
table(p3_data$alt_ra_1)
```

If someone responded `I prefer to not answer` or `I don't know`, we will recode with missing. 

```{r}
p3_data <- p3_data |>
  mutate(across(contains("alt_ra_"), ~na_if(., 8))) # I prefer to not answer
p3_data <- p3_data |>
  mutate(across(contains("alt_ra_"), ~na_if(., 7))) # I don't know
```

Let's see if it worked:

```{r}
get_value_labels(p3_data$alt_ra_1)
table(p3_data$alt_ra_1)
```
Great, it worked!

### Support from Alter

Likewise for this variable.

```{r}
get_value_labels(p3_data$alt_idsupp_1_1)
table(p3_data$alt_idsupp_1_1)
```

We need to recode `Does not apply` as missing.

```{r}
p3_data <- p3_data |>
  mutate(across(contains("alt_idsupp_"), ~na_if(., 7))) # Does not apply
```

Let's see if it worked.

```{r}
get_value_labels(p3_data$alt_idsupp_1_1)
table(p3_data$alt_idsupp_1_1)
```

## Create Ego Network Data

Let's create our ego network data! 
```{r}
#| eval: false
ego_object <- onefile_to_egor(
  egos = p3_data,
  ID.vars = list(ego = "egoid"),
  attr.start.col = "alt_rel_1", # Name of Variable with the first alter attribute
  attr.end.col = "alt_smv_conc_10", # Name of variable with last alter attribute
  max.alters = 10, # Number of maximum alters that were named by participants
  aa.first.var = "ap1_2") # Name of first alter-alter variable
```

The resulting output is an object that is comprised of 3 different datasets: EGO data (N = 953, Variables = 548), Alter data (N = 9,530, Variables = 145), and Alter-Alter Tie Data (N = 14,460; Variabels = 4).


As the warning suggested, we need to remove invalid weights. A weight is invalid if it is empty, which signifies missing data. As such, we will remove those values. 

```{r}
#| eval: false
ego_object <- ego_object |> 
  activate(aatie) |> 
  filter(weight != "")
```

Next, lets save our ego object. 
```{r}
#| eval: false
saveRDS(ego_object, here("work_data/egor_object.rds"))
```


```{r}
#| include: false
ego_object <- readRDS(file =  here("work_data/egor_object.rds"))
```

## Save Results
```{r}
saveRDS(p3_data, "work_data/p3_data.Rds")
```