---
title: "Ego-Network Variables"
format:
  html: 
    toc: true
    toc-depth: 3
    embed-resources: true
output: html_document
editor_options: 
  chunk_output_type: console
---
:::callout-note
**Overall Summary**: This phase of data cleaning will

- prepare data for creating ego-network datasets
- correct alter variable names
- correct alter values
:::
## Preparations
### Packages
```{r}
pacman::p_load(tidyverse, haven, here, labelled, egor, igraph)
```
## Import Dataset
```{r}
p3_data <- readRDS("work_data/p2_data.Rds")
```

## Prepare Data
### Create EgoID variable
```{r}
p3_data <- p3_data |>
  mutate(egoid = 1:n()) |>
  select(egoid, everything())
```

### Remove All Text Variables
EgoR package cannot create an ego network data if text variables/values exist. Thus if any variables contain text, we will drop.

```{r}
p3_data <- p3_data |> 
  select(-contains("TEXT"))
```

## Correct Variable Names

```{r}
p3_data <- p3_data %>%
  rename(
  # NEW var = OLD var
  alt_convs_2 = alt_cons_2,
  alt_ra_2 = alr_ra_2,
  alt_ra_3 = alr_ra_3
  )
```


## Correct Alter-Alter Ties 

The value labels are coded in a manner that will bias how we calculate ego network characteristics. 

```{r}
get_variable_labels(p3_data$ap1_2)
get_value_labels(p3_data$ap1_2)
```

As you can see, `No = 2` while `Yes = 1`. However, when we calculate the sum, we want to make sure that we only count the `Yes`'s, thus it makes more sense to code No as 0, and Yes as 1. 

```{r}
p3_data <- p3_data |>
  mutate(across(contains("ap1_") | contains("ap2_") | contains("ap3_") | contains("ap4_") | contains("ap5_") | contains("ap6_") | contains("ap7_") | contains("ap8_") | contains("ap9_"), ~ifelse(. == 2, 0, .)))
```

The code above specifies: `If value is equal to 2, then code as 0. If the condition is not true, leave as is`.

## Create Ego Network Data

Let's create our ego network data! 
```{r}
#| eval: false
ego_object <- onefile_to_egor(
  egos = p3_data,
  ID.vars = list(ego = "egoid"),
  attr.start.col = "alt_rel_1", # Name of Variable with the first alter attribute
  attr.end.col = "alt_smv_conc_10", # Name of variable with last alter attribute
  max.alters = 10, # Number of maximum alters that were named by participants
  aa.first.var = "ap1_2") # Name of first alter-alter variable


saveRDS(ego_object, here("work_data/egor_object.Rds"))
```

```{r}
#| include: FALSE
ego_object <- readRDS(here("work_data/egor_object.Rds"))
```

The resulting output is an object that is comprised of 3 different datasets: EGO data (N = 953, Variables = 548), Alter data (N = 9,530, Variables = 145), and Alter-Alter Tie Data (N = 14,460; Variabels = 4).

### Create Merged Dataset

Having created the three different datasets, we want to create a dataset that merges the ego and alter data. 

```{r}
p3.1_data <- ego_object$ego

p3.2_data <- ego_object$alter
```

What we've done above is create two datasets based on ego and alter (respectively). Now using the .egoID variable, lets merge these two datasets together. 

```{r}
p3.3_data <- p3.2_data %>%
  left_join(p3.1_data, by = ".egoID")
```

## Create Network Characteristic Variables

A weight is invalid if it is empty, which signifies missing data. As such, we will remove those values. This is also **essential** as a weight must range between 0 to 1 to calculate network characteristic variables (e.g., network density).

```{r}
ego_object <- ego_object |> 
  activate(aatie) |> 
  filter(weight != "")
```

### Network Density

```{r}
data_density <- ego_density(ego_object)
data_density$.egoID <- as.numeric(data_density$.egoID)
data_density$density <- as.numeric(data_density$density)
psych::describe(data_density$density)

data_density %>%
  ggplot(aes(x = density)) +
  geom_bar()
```

Here we can see that the average network density is 0.34 (SD = 0.26). Having created this variable, we want to merge with the larger dataset we are working with. 

```{r}
p3.3_data$.egoID <- as.numeric(p3.3_data$.egoID)
p3.4_data <- left_join(p3.3_data, data_density, by = ".egoID")
```

### Create Alter Sum 

Next, let's create a variable for each ego that counts how many alters they named. 

We do this by counting the distinct number of observations for each ego. 

```{r}
p3.3_data <- p3.3_data %>%
  group_by(.egoID) %>%
  mutate(alt_sum = n_distinct(.altID)) %>%
  ungroup()

psych::describe(p3.3_data$alt_sum)
```

Looks like everyone named 10 alters.

`alt_sum` was added to `p3.3_data` after `p3.4_data` was already created, so it needs to be explicitly merged in.

```{r}
p3.4_data <- left_join(p3.4_data, p3.3_data %>% distinct(.egoID, alt_sum), by = ".egoID")
```

## Create igraph Object

To create component variables, we need to convert the egoR object into an igraph object.

```{r}
#| eval: false
igraphs <- as_igraph(ego_object)
saveRDS(igraphs, here("work_data/igraphs.Rds"))
```

```{r}
#| include: FALSE
igraphs <- readRDS(here("work_data/igraphs.Rds"))
```

Here we can previous whan the igraph output looks like for Ego1.
```{r}
print(igraphs[[1]])
```

### Network Components (Fragmentation)

`components()` identifies clusters of alters who are connected to each other, independent of ego. An ego whose alters split into several components has a *fragmented* network — the alters don't all know one another, so ego sits at the center of several disconnected social worlds. An ego with one component has a fully reachable network, at least in terms of connectivity (this doesn't necessarily mean the network is *dense*).

We calculate, for each ego: 
-  the number of components, and 
-  the proportion of alters who fall in the largest component (a common fragmentation index — closer to 1 means most alters are reachable from one another; lower values indicate a more splintered network).

```{r}
data_components <- tibble(
  .egoID = sapply(igraphs, function(g) graph_attr(g, ".egoID")),
  n_components = sapply(igraphs, function(g) components(g)$no),
  largest_component_size = sapply(igraphs, function(g) max(components(g)$csize)),
  prop_largest_component = sapply(igraphs, function(g)
    max(components(g)$csize) / vcount(g))
)

data_components$.egoID <- as.numeric(data_components$.egoID)
psych::describe(data_components[, c("n_components", "prop_largest_component")])

data_components %>%
  ggplot(aes(x = n_components)) +
  geom_bar()
```

Having created these variables, we merge them into the larger dataset.

```{r}
p3.4_data <- left_join(p3.4_data, data_components, by = ".egoID")
```

### Network Centrality

Centrality measures describe *positions of individual alters* within an ego's network, rather than the network as a whole. To turn them into ego-level variables, we take two approaches: (1) summarize the alter-level scores (mean, max) across each ego's alters, and (2) use igraph's *centralization* functions, which collapse alter-level centrality into a single Freeman-style graph index — how dominated the network is by one or a few central alters (0 = evenly distributed, 1 = a pure hub/star structure).

**Betweenness** identifies "broker" alters — those who sit on the shortest path between two other alters. A network with high betweenness centralization is one where a single alter (or few) bridges otherwise-separate parts of ego's social world.

**Edge betweenness** applies the same logic to ties rather than alters — it flags which alter-alter connections are critical bridges holding the network together (removing them would disconnect otherwise-linked parts of the network).

**Closeness** captures how quickly each alter can be reached from every other alter on average. High closeness centralization means the network revolves around one or a few well-embedded alters, rather than everyone being roughly equidistant from one another.

```{r}
#| warning: false
data_centrality <- tibble(
  .egoID = sapply(igraphs, function(g) graph_attr(g, ".egoID")),

  # Alter-level betweenness, summarized to the ego level
  mean_betweenness = sapply(igraphs, function(g) mean(betweenness(g), na.rm = TRUE)),
  max_betweenness  = sapply(igraphs, function(g) max(betweenness(g), na.rm = TRUE)),

  # Edge-level betweenness, summarized to the ego level
  mean_edge_betweenness = sapply(igraphs, function(g) mean(edge_betweenness(g), na.rm = TRUE)),

  # Alter-level closeness, summarized to the ego level
  mean_closeness = sapply(igraphs, function(g) mean(closeness(g), na.rm = TRUE)),

  # Graph-level centralization indices (Freeman-style)
  centralization_betweenness = sapply(igraphs, function(g) centr_betw(g)$centralization),
  centralization_closeness   = sapply(igraphs, function(g) centr_clo(g)$centralization)
)

data_centrality$.egoID <- as.numeric(data_centrality$.egoID)
psych::describe(data_centrality[, c("centralization_betweenness", "centralization_closeness")])

data_centrality %>%
  ggplot(aes(x = centralization_betweenness)) +
  geom_bar()
```

Having created these variables, we merge them into the larger dataset.

```{r}
p3.4_data <- left_join(p3.4_data, data_centrality, by = ".egoID")
```

## Create Separate Datasets

Now that all ego-level network characteristics have been merged in, let's split the data into the three datasets we actually want to work with: ego data (one row per ego), ego-alter data (one row per alter, with ego attributes attached), and alter-alter data (the tie list).

### Ego Data

```{r}
p3_ego_data <- p3.1_data %>%
  mutate(.egoID = as.numeric(.egoID)) %>%
  left_join(data_density, by = ".egoID") %>%
  left_join(data_components, by = ".egoID") %>%
  left_join(data_centrality, by = ".egoID") %>%
  left_join(p3.3_data %>% distinct(.egoID, alt_sum), by = ".egoID")
```

### Ego-Alter Data

`p3.4_data` already *is* our ego-alter dataset — one row per alter, with the ego's survey responses and all ego-level network characteristics attached to every alter row.

```{r}
p3_ego_alter_data <- p3.4_data
```

### Alter-Alter Data

```{r}
p3_alter_alter_data <- ego_object$aatie
```

## Save Results
```{r}
saveRDS(p3.4_data, "work_data/p3_data.Rds")
```

### Codebooks

We will generate a .txt codebook for the ego data and ego-alter data (the alter-alter tie list is just IDs and a weight, so it doesn't need one).

```{r}
#| eval: FALSE
memisc::Write(memisc::codebook(p3_ego_data),
      file = "output_txt/p3_ego_data_codebook.txt")

memisc::Write(memisc::codebook(p3_ego_alter_data),
      file = "output_txt/p3_ego_alter_data_codebook.txt")

writexl::write_xlsx(p3_ego_data, "work_data/p3_ego_data.xlsx")
writexl::write_xlsx(p3_ego_alter_data, "work_data/p3_ego_alter_data.xlsx")
writexl::write_xlsx(p3_alter_alter_data, "work_data/p3_alter_alter_data.xlsx")
```